---
title: "R Notebook"
output: html_notebook
---
# Libraries
```{r}
library(tidyverse)
library(janitor)
library(lubridate)
library(GGally)
library(broom)
library(pROC)
library(modelr)
library(caret)
library(gghighlight)
library(here)
```

# Reading in cleaned data
```{r}
loans_cleaned <- read_csv("../data/loans_cleaned.csv")
```

# Splitting data into background_check data 
```{r}
# selecting vars used during the background checks of customer
background_checks <- loans_cleaned %>% 
  select(c(defaulted,
           verification_status, # category - verified income
           dti, # dbl - LC ratio on monthly debts/monthly income
           delinq_2yrs,# dbl - The number of 30+  delinquency in the borrower's credit file for the past 2 years
           earliest_cr_line, # dbl - The month the borrower's earliest reported credit line was opened
           fico_range_low, # dbl - credit score at loan origin
           inq_last_6mths, # dbl - The number of inquiries in past 6 months (excluding auto and mortgage inquiries)
           mths_since_last_delinq, # dbl - The number of months since the borrower's last delinquency.(lc??)
           mths_since_last_record, # The number of months since the last public record.
           open_acc, # dbl - The number of open credit lines in the borrower's credit file.
           pub_rec, # dbl - The number of open credit lines in the borrower's credit file.
           revol_bal, # dbl - Total credit revolving balance  unpaid credit at the end of a billing cycle
           revol_util, # chr - %, amount of credit the borrower is using relative to all available revolving credit
           total_acc, # dbl - The total number of credit lines currently in the borrower's credit file
           delinq_amnt, # dbl - The past-due amount owed for the accounts on which the borrower is now delinquent, 
           pub_rec_bankruptcies,
           fico_range_low,
           fico_range_high
  )) 

```

# Splitting data into onramp_data
```{r}
# selecting vars used during on ramp phase of delivering loans
# not including funded/inv due to no use
# using grade rather than sub for correlation affect
onramp_data  <- loans_cleaned %>% 
  select(c(defaulted,
           loan_amnt,
           term,
           int_rate,
           installment,
           grade, # LC assigned loan subgrade, can use grade
           emp_title, # chr - not really useful
           emp_length, # chr - years, should turn to factor and use as bins/dummy
           home_ownership, # chr - binned
           issue_d,
           loan_status,
           pymnt_plan, # chr - turn logical
           desc, # chr - text, no use
           purpose, # chr - binned - turn to factor
           title, # chr - user text to describe loan
           zip_code, # chr - usefull for binning, redacted
           state_name, # chr - turn to factor, no need for abbreviated var
           fico_range_low,
           fico_range_high, 
  )) 

# -----------------------------------------------------------------------------#

# removing user entered text data & loan_status as created flag, & pymnt_plan as all n bar 1
# removing fico_range_low to have a higher tolerance for credit rate
# removing application_type as they are all individual

# onramp_data_trimmed  <- onramp_data_trimmed %>% 
#  select(-c(emp_title, desc, title, loan_status, pymnt_plan, fico_range_low, application_type))

# -----------------------------------------------------------------------------#
```

#------------------------------------------------------------------------------#

# Stats
### Percentage of LC customers classified as defaulting RESULTS - DEFAULTS
```{r}
# about 15% of customers are either charged off, late > 16days or have defaulted
loans_cleaned %>% 
  group_by(defaulted) %>% 
  summarise(Count = n(),
            Percent = round((Count/nrow(loans_cleaned))* 100, 2)) %>% 
  knitr::kable( 
             caption = "Percentage of LendingClub customers classified as defaulting",
             col.names = c("Defaulted", 
                           "Count",
                           "%"),
             "pipe")
```

### Mean/median number of loans per state + Ratio of defaulted per state RESULTS - GEOGRAPHY
```{r}
# total ratio of defaulted per state
default_rate_per_state <- loans_cleaned %>%
  group_by(state_name) %>%
  summarise(count = n(), defaulted_rate = sum(defaulted == "Yes") / n()) %>% 
  arrange(desc(defaulted_rate))

# Top 5 states with highest default rate
default_rate_per_state %>% 
  summarise(state_name, count, defaulted_rate = round(defaulted_rate*100, 2)) %>% 
  slice(1:5) %>% 
  knitr::kable( 
             caption = "Top 5 states with highest default rate",
             col.names = c("State", 
                           "Number of defaulted loans",
                           "Defaulted Rate, %"),
             "pipe")
  
# 847.4
default_rate_per_state %>%
  summarise(mean(count))

# 463.5
default_rate_per_state %>%
  summarise(median(count))
```

#------------------------------------------------------------------------------#

# Graphs
### Graph Theme
```{r}
lc_theme <- function(){
  theme(panel.background = element_rect(fill = "ghostwhite"),
        legend.key = element_rect(fill = "#0047b3"),
        panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "none",
        axis.title = element_text(face = "bold"),
        plot.title = element_text(face = "bold")
  ) +
    theme_minimal(base_size = 14)
}
# dodgerblue4
# firebrick3
# ghostwhite
```

### Map showing rate of defaults across USA RESULTS - GEOGRAPHY
```{r}
# str to lower to match mapping library
# creating map
default_rate_per_state %>% 
  select(state_name, defaulted_rate) %>%
  rename(region = state_name, value = defaulted_rate) %>% 
  mutate(region = str_to_lower(region)) %>% 
  choroplethr::state_choropleth(title = "Default Rate Per State")
```

### Barchart - Total LendingClub Loans Per State RESULTS - GEOGRAPHY
```{r}
# proportional to state size
# note, cal has 1	California	39,538,223
# 2	Texas	- 29,145,505
# 3	Florida	- 21,538,187
# 4	New York	- 20,201,249

# total LC loans by state
top5_states <- loans_cleaned %>% 
  count(state_name, sort = T) %>% 
  slice(1:5) 

loans_cleaned %>% 
  count(state_name) %>% 
  ggplot() +
  aes(x = n, y = reorder(state_name, n)) +
  geom_col(fill = "dodgerblue4") +
  lc_theme() +
  labs( 
    title = "Total LendingClub Loans Per State",
    subtitle = "Between 2007 - 2011",
    x = "Loans",
    y = "States") +
  gghighlight(n >= min(top5_states$n))
```

### Barchart - Total Defaulted LendingClub Loans Per State RESULTS - GEOGRAPHY
```{r}
# total defaulted loans by state 
top5_defaulted_states <- loans_cleaned %>% 
  filter(defaulted == "Yes") %>% 
  count(state_name, sort = T) %>% 
  slice(1:5) 

loans_cleaned %>% 
  filter(defaulted == "Yes") %>% 
  count(state_name, sort = T) %>% 
  ggplot() +
  aes(x = n, y = reorder(state_name, n)) +
  geom_col(fill = "firebrick3") +
  lc_theme() +
  labs( 
    title = "Total Defaulted LendingClub Loans Per State",
    subtitle = "Between 2007 - 2011",
    x = "Loans",
    y = "States") +
  gghighlight(n >= min(top5_defaulted_states$n))
```

### Barchart - Ratio of Defaulted LendingClub Loans Per State(highlighting top 5 states above avg loan count) RESULTS - GEOGRAPHY
```{r}
# taking top 5 states with highest default rate with a count more than the average amnt of loans per state
# mean number of loans per state 847.4	
top5_defaulted_states_mean_loans <- default_rate_per_state %>% 
  filter(count > mean(count)) %>% 
  arrange(desc(defaulted_rate)) %>%
  slice(1:5) 

# Top 5 states above average loan count with highest default rate
default_rate_per_state %>% 
  filter(count > mean(count)) %>% 
  arrange(desc(defaulted_rate)) %>%
  summarise(State = state_name, 
            "Number of defaulted loans" = count,
            "Defaulted Rate, %" = round(defaulted_rate*100, 2)) %>% 
  slice(1:5) %>% 
   knitr::kable(caption = "Top 5 states above average loan count with highest default rate")

# Top 5 states with highest default rate
# shows Nebraska has highest default rate although small proportion
default_rate_per_state %>% 
  arrange(desc(defaulted_rate)) %>%
  summarise(State = state_name, 
            "Number of defaulted loans" = count,
            "Defaulted Rate, %" = round(defaulted_rate*100, 2)) %>% 
  slice(1:5) %>% 
   knitr::kable(caption = "Top 5 states with highest default rate")


# States with lowest default rate that are above average loan count, 'Least likely to default on loans
default_rate_per_state %>% 
  filter(defaulted_rate < mean(defaulted_rate) & count > mean(count)) %>% 
  arrange(defaulted_rate) %>%
  summarise(State = state_name, 
            "Number of defaulted loans" = count,
            "Defaulted Rate, %" = round(defaulted_rate*100, 2)) %>% 
 # slice(26:31) %>% 
   knitr::kable(caption = "States with lowest default rate that are above average loan count, 'Least likely to default on loans'")
  
# states with highest default rate plot
default_rate_per_state %>% 
  ggplot() +
  aes(x = defaulted_rate, y = reorder(state_name, defaulted_rate)) +
  geom_col(fill = "grey82") +
  scale_x_continuous(limits =c (0, 1)) +
  lc_theme() +
  labs( 
    title = "Ratio of Defaulted LendingClub Loans Per State",
    subtitle = "Between 2007 - 2011",
    x = "Loans",
    y = "States") +
  geom_col(data = top5_defaulted_states_mean_loans, fill = "firebrick3")
```

#------------------------------------------------------------------------------#

### Barchart - Volume of LendingClub Loan Catagories DATA
```{r}
# showing types of loan status
loans_cleaned %>% 
  group_by(loan_status) %>% 
  count(loan_status, sort = T) 

# looking at volume within loan status categories
loans_cleaned %>% 
  count(loan_status) %>% 
  ggplot() +
  aes(y = reorder(loan_status, n),  x = n) +
  geom_col(fill = "dodgerblue4") +
  lc_theme() +
  labs( 
    title = "Volume of LendingClub Loan Catagories",
    subtitle = "Between 2007 - 2011",
    x = "",
    y = "") +
  geom_text(aes(label = n), hjust = -0.1, vjust = 0) +
  theme(axis.text.y = element_text(colour  = c('firebrick3',
                                            'firebrick3', 
                                            'firebrick3',
                                            'grey50',
                                            'grey50',
                                            'firebrick3',
                                            'grey50',
                                            'firebrick3',
                                            'grey50'),
                                   face = "bold")) 
```

### Types of LendingClub Loans, 2007 - 2011 DATA 
```{r}
#-------------------------------
# data from mid 2007 - end 2011

# total volume of LC loans over time
loans_cleaned %>% 
  count(issue_d, sort = T) %>% 
  ggplot() +
  aes(x = issue_d, y = n) +
  geom_line()

# counting purpose of loan types
loans_cleaned %>% 
  count(purpose, sort = T) %>% 
  summarise(Purpose = purpose, Count = n, 
            Percent = round((Count/nrow(loans_cleaned))* 100, 2)) %>% 
  knitr::kable()


# types of p2p loans over time
loans_cleaned %>% 
  group_by(purpose, issue_d) %>% 
  summarise(count = n()) %>% 
  ggplot() +
  aes(x = issue_d, y = count, colour = purpose) +
  geom_line() +
  lc_theme() +
  labs( 
    title = "Types of LendingClub Loans",
    subtitle = "Between 2007 - 2011",
    x = "",
    y = "Volume",
    colour = "Purpose")
#---------------------------
```

#------------------------------------------------------------------------------#

### Barplot showing volume of negative vs positive loan status RESULTS - PURPOSE
```{r}
# plot showing volume of negative vs positive loan status ratio no/yes
loans_cleaned %>%
  group_by(purpose, defaulted) %>% 
  count(purpose) %>% 
  ggplot() +
  aes(y = reorder(purpose, n), 
      x = n, 
      fill = defaulted) +
  geom_col() +
  labs( 
    title = "Volume of defaulted to non-defaulted loans per purpose",
    subtitle = "Between 2007 - 2011",
    x = "Loans",
    y = "Purpose",
    fill = "Defaulted") +
  scale_fill_discrete(limits = c("Yes", "No")) +
  theme(axis.text.y = element_text(colour  = c('firebrick3',
                                               'grey50',
                                               'firebrick3',
                                               'grey50',
                                               'grey50',
                                               'grey50',
                                               'dodgerblue4',
                                               'dodgerblue4',
                                               'firebrick3',
                                               'dodgerblue4',
                                               'dodgerblue4',
                                               'grey50',
                                               'dodgerblue4',
                                               'grey50'),
                                   face = "bold")) 

# plot showing ratio of negative vs positive loan status ratio 
loans_cleaned %>%
  mutate(purpose = forcats::fct_reorder(.f = purpose, 
                                         .x = defaulted,
                                         .fun = function(.x) mean(.x == "Yes"),
                                         .desc = F)) %>%
  ggplot(aes(y = purpose, fill = defaulted)) +
  geom_bar(position = "fill") +
  lc_theme() +
  labs( 
    title = "Ratio of defaulted to non-defualted loans per purpose",
    subtitle = "Between 2007 - 2011",
    x = "Ratio",
    y = "Purpose",
    fill = "Defaulted") +
  scale_fill_discrete(limits = c("Yes", "No"))

# table for purpose volumes of defaults 
loans_cleaned %>% 
  filter(defaulted == "Yes") %>% 
  count(purpose, sort = T) %>% 
  knitr::kable( 
    caption = "Volume of defaults per purpose",
    col.names = c("Purpose", 
                  "Number of Defaults"),
    "pipe")

# default rates per purpose
default_rate_per_purpose <- loans_cleaned %>%
  group_by(purpose) %>%
  summarise(count = n(), defaulted_rate = sum(defaulted == "Yes") / n()) %>% 
  arrange(desc(defaulted_rate))

# Top 5 purpose with highest default rate
default_rate_per_purpose %>% 
  #filter(count > mean(count)) %>% 
  summarise(purpose, 
            count, 
            defaulted_rate = round(defaulted_rate*100, 2)) %>% 
  arrange(desc(defaulted_rate))
  #slice(1:5)

# shows that other category is a mixture of repaying debts, credit cards, repairs, building credit, for family members /friends
loans_cleaned %>% 
  filter(purpose == "other") %>% 
  summarise(desc)
```

#------------------------------------------------------------------------------#

### Linegraph - Fico_range_low RESULTS - FICO RANGE 
```{r}
# as the lower end of the fico range increases so does the amount of defaulted loans,
# it shows an uneven bell curve being skewed to the right, 65 - 685 fico range low has the highest amount of defaults, then drops off in terms of number of defaults
loans_cleaned %>% 
  group_by(fico_range_low, defaulted) %>% 
  filter(defaulted == "No") %>% 
  count(fico_range_low, sort = T) %>% 
  arrange(fico_range_low)

# FICO score line graph
loans_cleaned %>% 
  group_by(fico_range_low, defaulted) %>% 
  count(fico_range_low, sort = T) %>% 
  ggplot() +
  aes(x = fico_range_low, y = n, colour = defaulted) +
  geom_line() +
    lc_theme() +
  labs( 
    title = "FICO Score of customers defaulting on loans ",
    subtitle = "Between 2007 - 2011",
    x = "FICO score",
    y = "Volume of customers",
    colour = "Defaulted") +
  scale_colour_discrete(limits = c("Yes", "No"))+
  scale_x_continuous(breaks=seq(600, 830,10))

# FICO sratio bar chart
loans_cleaned %>%
  ggplot(aes(x = fico_range_low, fill = defaulted)) +
  geom_bar(position = "fill") +
      lc_theme() +
  labs( 
    title = "Ratio of FICO Scores defaulting on loans ",
    subtitle = "Between 2007 - 2011",
    x = "FICO score",
    y = "Ration of defaults",
    fill = "Defaulted") +
  scale_colour_discrete(limits = c("Yes", "No")) +
  scale_x_continuous(breaks=seq(600, 830,10)) +
  theme(axis.text.x = element_text(angle=45))
```

# LogReg Model
```{r}
library(caret)
library(GGally)

# correlation matrix
loans_trim %>% 
  select(c(inq_last_6mths, loan_amnt, loan_status_flag, fico_range_low, fico_range_high)) %>% 
  ggcorr(label = T, layout.exp = 1)

# scaling data 
loans_data_scaled <- as.data.frame(scale(loan_trim))

mod1 <- glm(loan_status_flag ~ inq_last_6mths,
            data = loan_trim,
            family = binomial((link = "logit")))

summary(mod1)
```

### Setting up correlations
```{r}
# splitting df down to compare easily 
split1 <- onramp_data_trimmed %>% 
  select(defaulted, loan_amnt, term, int_rate, installment)

split2 <-  onramp_data_trimmed %>% 
  select(defaulted, grade, emp_length, home_ownership, annual_inc)

split3<- onramp_data_trimmed %>% 
  select(defaulted, issue_d, purpose, zip_code, state_name, fico_range_high)
```

### ggpairs()
```{r}
# defaulted + term(cat) > int_rate > loan_amnt 
split1 %>% 
  ggpairs()

# defaulted + annual_inc(cont) > home_ownership > emp_length
# b-grades are most likely to default
# lower income customers are more likely to default
split2 %>% 
  ggpairs()

# defaulted + fico_range_high(cat) > purpose (cat)
split3 %>% 
  ggpairs(cardinality_threshold = NULL)
```

### Creating models
```{r}
# categorical
mod1_1pred_term <- glm(defaulted ~ term, 
                                  data = onramp_data_trimmed, 
                                  family = binomial(link = 'logit'))

# continuous
mod2_1pred_annual_inc <- glm(defaulted ~ annual_inc, 
                                  data = onramp_data_trimmed, 
                                  family = binomial(link = 'logit'))

# categorical
mod3_1pred_purpose <- glm(defaulted ~ purpose, 
                                  data = onramp_data_trimmed, 
                                  family = binomial(link = 'logit'))

# combined model
mod5_1pred_combined <- glm(defaulted ~ .,
                                  data = onramp_data_trimmed, 
                                  family = binomial(link = 'logit'))
clean_names(tidy(mod5_1pred_combined))

# creating null
null_model <- glm(defaulted ~ 1,
                  data = onramp_data_trimmed,
                  family = binomial(link = 'logit'))

anova(mod5_1pred_combined, null_model, test = "Chisq")


# all p-values are below 0, so tests are significant 
clean_names(tidy(mod1_1pred_term))
clean_names(tidy(mod2_1pred_annual_inc))
clean_names(tidy(mod3_1pred_purpose))

clean_names(glance(mod1_1pred_term))
clean_names(glance(mod2_1pred_annual_inc))
clean_names(glance(mod3_1pred_purpose))
```

### ROC curves
```{r}
onramp_data_trimmed_with_mod1 <- onramp_data_trimmed %>%
  add_predictions(mod1_1pred_term, type = "response")

onramp_data_trimmed_with_mod2 <- onramp_data_trimmed %>%
  add_predictions(mod2_1pred_annual_inc, type = "response")

onramp_data_trimmed_with_mod3 <- onramp_data_trimmed %>%
  add_predictions(mod3_1pred_purpose, type = "response")

roc_obj_mod1 <- onramp_data_trimmed_with_mod1  %>%
  roc(response = defaulted, predictor = pred)

roc_obj_mod2 <- onramp_data_trimmed_with_mod2  %>%
  roc(response = defaulted, predictor = pred)

roc_obj_mod3 <- onramp_data_trimmed_with_mod3  %>%
  roc(response = defaulted, predictor = pred)

roc_curve <- ggroc(
  data = list(
    mod1 = roc_obj_mod1, 
    mod2 = roc_obj_mod2,
    mod3 = roc_obj_mod3
  ), 
  legacy.axes = TRUE) +
  coord_fixed()

roc_curve

# mod1 using term of loan seems out the bunch to be the best classifier as its "curve" is closest to the top left corner. Not really over the diagonal line for a random classifier for threshold values though...
```

```{r}
auc(roc_obj_mod1)
auc(roc_obj_mod2)
auc(roc_obj_mod3)
```

### Cross validation, using a test and a train set
```{r}
train_control <- trainControl(method = "repeatedcv", 
                              number = 5,
                              repeats = 100,
                              savePredictions = TRUE, 
                              classProbs = TRUE, 
                              summaryFunction = twoClassSummary)
```

```{r}
mod1_1pred_term_cv <- train(mod1_1pred_term$formula,
                                       data = onramp_data_trimmed,
                                       trControl = train_control,
                                       method = "glm",
                                       family = binomial(link = 'logit'))
```

```{r}
mod2_1pred_annual_inc_cv <- train(mod2_1pred_annual_inc$formula,
                              data = onramp_data_trimmed,
                              trControl = train_control,
                              method = "glm",
                              family = binomial(link = 'logit'))
```

```{r}
mod3_1pred_purpose_cv <- train(mod3_1pred_purpose$formula,
                           data = onramp_data_trimmed,
                           trControl = train_control,
                           method = "glm",
                           family = binomial(link = 'logit'))
```

```{r}
# near enough same AUC as when used on full data set - with one classifier hard to overfit
mod1_1pred_term_cv$results
mod2_1pred_annual_inc_cv$results
mod3_1pred_purpose_cv$results
```

### Interpretation of the fitted coefficient with odds ratio
```{r}
coeff <- clean_names(tidy(mod1_1pred_term)) %>%
  filter(term == "term") %>%
  select(estimate)
coeff
```

```{r}
# this shows the odd of defaulting on a loan are increased by multiplication by a factor of 1.03
odds_ratio <- exp(coeff * 1)
odds_ratio
```

# Log Reg with all potential predictors
```{r}
mod4_all_preds <- glm(defaulted ~ ., 
                      data = onramp_data_trimmed, 
                      family = binomial(link = 'logit'))

clean_names(tidy(mod4_all_preds)) %>%
  select(term, p_value) %>%
  filter(p_value > 0.05)
```

```{r}
onramp_data_trimmed_with_mod4 <- onramp_data_trimmed %>%
  add_predictions(mod4_all_preds, type = "response")

roc_obj_mod4 <- onramp_data_trimmed_with_mod4  %>%
  roc(response = defaulted, predictor = pred)

ggroc(roc_obj_mod4) +
  coord_fixed()

# AUC and ROC curve could be better..
auc(roc_obj_mod4)
```

```{r}
# will not run
mod4_all_preds_cv <- train(defaulted ~ .,
                           data = onramp_data_trimmed,
                           trControl = train_control,
                           method = "glm",
                           family = binomial(link = 'logit'))
```


